### 2013-12-22 manuscript/AGU-LaTeX/ms_main.tex
v1 of figures, v0.2 of manuscript

@@

This parent model was used before to investigate the consequences of increased plant mortality. It's simple formulation of additional plant stress can be interpreted as the effects of increased grazing intensity by large herbivores (K\'efi et al 2007a). However, the model has two features that are unrealistic: the individual plant mortality which is caused by grazing (1) did not vary with total vegetation cover, and (2) it applied globally, affecting all plants likewise (Fig. 1).
Both assumptions are justified by the need for model simplicity, but disallow a translation of the model parameters to real livestock rates. The relationship between livestock rates and the stability of vegetation cover is a crucial information for management decisions. The present study aims at increasing the realism and applicability of the model by applying two extensions that enable simulations of a grazing gradient on drylands.

%Grazing depends on vegetation cover
The first extension adds a contrasting assumption on the dependence on the total vegetation cover to the parent model. The prior model defined grazing as a constant risk of additional plant death. On the population level, the number of plants dying because of grazing increased with vegetation cover (K\'efi et al. 2007, Fig. 1, black solid lines). However, grazers forage in the landscape and consume to fulfill their daily demands. In other words, the number of plants suffering from grazing is rather constant with vegetation cover. As a consequence, at low vegetation cover the grazing risk on individual plants will be high, whereas when vegetation cover is high, risk will be shared among many plants.
In reality, the total feeding of large herbivores in the landscape is not entirely independent of vegetation cover but follows a non-linear saturating curve (type-II functional response, Holling 1959, Spalinger and Hobbs 1992). Thus, particularly at low vegetation cover, grazing is limited by the search success of the grazer, which is resource-density dependent. In rangeland systems, however, the functional response at low vegetation cover depends on factors on a larger scale: the animals move between landscapes motivated by internal decision patterns or their density is modulated by human management decisions (Senft 1989). We assume therefore, that smooth type-II functional responses, as described for wild ungulates, are unlikely to happen in intensive rangelands within the spatial scale adressed in the present study.

>>>>

\begin{figure}[t]
	\centering
		\includegraphics{figures/fig1.pdf}
	\caption{a) Vegetation in arid ecosystems grows in patches. The associated growth provides enhanced local growth conditions as well as protection against grazing. With decreasing environmental quality, vegetation declines and the landscape becomes more fragmented (mid column), leading to desertification (right column). b) Schematic representation of plants growing in patches. The direct neighborhood of plants provides shelter, retention of organic matter and water below and atop the soil surface. Spots remote from vegetation suffer erosion and evaporation. c) The cellular automata model used in this study distinguishes three states: occupied by vegetation (black), empty but fertile (grey) and degraded (white). d) Cumulative patch size distributions described by power law models: with decreasing environmental quality, the number of large patches declines, turning the power law distribution from up-bent (left) to down-bent (mid). Degraded landscapes have lost all the larger patches and are not to be described by a power law (right). }
	\label{fig:fig1}
\end{figure}

@@

This parent model was used before to investigate the consequences of increased plant mortality. It's simple formulation of additional plant stress can be interpreted as the effects of increased grazing intensity by large herbivores (K\'efi et al 2007a). However, the model has two features that are unrealistic: the individual plant mortality which is caused by grazing (1) did not vary with total vegetation cover, and (2) it applied globally, affecting all plants likewise (Fig. 1).

\begin{figure}[tp]
	\centering
		\includegraphics{figures/fig2.pdf}
	\caption{Grazing affects the mortality of individual plants and the patch structure emerging in the landscape. The four models compared in this study use different assumptions on the quantity and spatial differentiation of a particular grazing pressure. Visualized in the left column are the plants dying due to grazing (red). In each model, plants suffer different individual risk, sometimes depending on the number of direct local neighbors (estimates for plants with 4, 2, or 0 neighbors).  a) The parent model as formulated in Kéfi et al 2008 assumes that grazing affects a proportion of living plants, i.e. individual risk is independent of global vegetation cover and shared evenly among all plants. b) The livestock model assumes a constant number of plants dying, regardless of the number of plants available. The individual risk is increasing when vegetation cover is low. c) Assuming asso­cia­tive protection leaves plants in the patch center unaffected by grazing (4), where­as plants at the patch border (2) or growing isolated (0) are suffering more from grazing. Particularly at high vegetation cover, the proportional grazing concentrates on those plants. d) The combined model is integrating both assumptions leading to highest individual risk at low and high vegetation cover for isolated plants.  }
	\label{fig:fig1}
\end{figure}

Both assumptions are justified by the need for model simplicity, but disallow a translation of the model parameters to real livestock rates. The relationship between livestock rates and the stability of vegetation cover is a crucial information for management decisions. The present study aims at increasing the realism and applicability of the model by applying two extensions that enable simulations of a grazing gradient on drylands.

%Grazing depends on vegetation cover
The first extension adds a contrasting assumption on the dependence on the total vegetation cover to the parent model. The prior model defined grazing as a constant risk of additional plant death. On the population level, the number of plants dying because of grazing increased with vegetation cover (K\'efi et al. 2007, Fig. 2a). However, grazers forage in the landscape and consume to fulfill their daily demands. In other words, the number of plants suffering from grazing is rather constant with vegetation cover. As a consequence, at low vegetation cover the grazing risk on individual plants will be high, whereas when vegetation cover is high, risk will be shared among many plants.
In reality, the total feeding of large herbivores in the landscape is not entirely independent of vegetation cover but follows a non-linear saturating curve (type-II functional response, Holling 1959, Spalinger and Hobbs 1992). Thus, particularly at low vegetation cover, grazing is limited by the search success of the grazer, which is resource-density dependent. In rangeland systems, however, the functional response at low vegetation cover depends on factors on a larger scale: the animals move between landscapes motivated by internal decision patterns or their density is modulated by human management decisions (Senft 1989). We assume therefore, that smooth type-II functional responses, as described for wild ungulates, are unlikely to happen in intensive rangelands within the spatial scale adressed in the present study.

### 2013-12-22 manuscript/ms_main.tex
v1 of figures, v0.2 of manuscript

\begin{figure}[t]
	\centering
		\includegraphics{figures/fig1.pdf}
	\caption{a) Vegetation in arid ecosystems grows in patches. The associated growth provides enhanced local growth conditions as well as protection against grazing. With decreasing environmental quality, vegetation declines and the landscape becomes more fragmented (mid column), leading to desertification (right column). b) Schematic representation of plants growing in patches. The direct neighborhood of plants provides shelter, retention of organic matter and water below and atop the soil surface. Spots remote from vegetation suffer erosion and evaporation. c) The cellular automata model used in this study distinguishes three states: occupied by vegetation (black), empty but fertile (grey) and degraded (white). d) Cumulative patch size distributions described by power law models: with decreasing environmental quality, the number of large patches declines, turning the power law distribution from up-bent (left) to down-bent (mid). Degraded landscapes have lost all the larger patches and are not to be described by a power law (right). }
	\label{fig:fig1}
\end{figure}

For arid ecosystems, the major mechanism behind this bistability was found to be local facilitation of plant individuals (REF?). The presence of a plant provides its direct local environment with abilities of water retention, organic matter accumulation and protection against external stressors (REF?). As a consequence, the vegetation grows in patches (Aguiar and Sala 1999).  When implemented in a simple spatially explicit model, this positive neighborhood effect provides true bistability on the landscape scale (K\'efi et al 2007a,b).

This parent model was used before to investigate the consequences of increased plant mortality. It's simple formulation of additional plant stress can be interpreted as the effects of increased grazing intensity by large herbivores (K\'efi et al 2007a). However, the model has two features that are unrealistic: the individual plant mortality which is caused by grazing (1) did not vary with total vegetation cover, and (2) it applied globally, affecting all plants likewise (Fig. 1).

\begin{figure}[tp]
	\centering
		\includegraphics{figures/fig2.pdf}
	\caption{Grazing affects the mortality of individual plants and the patch structure emerging in the landscape. The four models compared in this study use different assumptions on the quantity and spatial differentiation of a particular grazing pressure. Visualized in the left column are the plants dying due to grazing (red). In each model, plants suffer different individual risk, sometimes depending on the number of direct local neighbors (estimates for plants with 4, 2, or 0 neighbors).  a) The parent model as formulated in Kéfi et al 2008 assumes that grazing affects a proportion of living plants, i.e. individual risk is independent of global vegetation cover and shared evenly among all plants. b) The livestock model assumes a constant number of plants dying, regardless of the number of plants available. The individual risk is increasing when vegetation cover is low. c) Assuming asso­cia­tive protection leaves plants in the patch center unaffected by grazing (4), where­as plants at the patch border (2) or growing isolated (0) are suffering more from grazing. Particularly at high vegetation cover, the proportional grazing concentrates on those plants. d) The combined model is integrating both assumptions leading to highest individual risk at low and high vegetation cover for isolated plants.  }
	\label{fig:fig1}
\end{figure}

Both assumptions are justified by the need for model simplicity, but disallow a translation of the model parameters to real livestock rates. The relationship between livestock rates and the stability of vegetation cover is a crucial information for management decisions. The present study aims at increasing the realism and applicability of the model by applying two extensions that enable simulations of a grazing gradient on drylands.

%Grazing depends on vegetation cover
The first extension adds a contrasting assumption on the dependence on the total vegetation cover to the parent model. The prior model defined grazing as a constant risk of additional plant death. On the population level, the number of plants dying because of grazing increased with vegetation cover (K\'efi et al. 2007, Fig. 2a). However, grazers forage in the landscape and consume to fulfill their daily demands. In other words, the number of plants suffering from grazing is rather constant with vegetation cover. As a consequence, at low vegetation cover the grazing risk on individual plants will be high, whereas when vegetation cover is high, risk will be shared among many plants.
In reality, the total feeding of large herbivores in the landscape is not entirely independent of vegetation cover but follows a non-linear saturating curve (type-II functional response, Holling 1959, Spalinger and Hobbs 1992). Thus, particularly at low vegetation cover, grazing is limited by the search success of the grazer, which is resource-density dependent. In rangeland systems, however, the functional response at low vegetation cover depends on factors on a larger scale: the animals move between landscapes motivated by internal decision patterns or their density is modulated by human management decisions (Senft 1989). We assume therefore, that smooth type-II functional responses, as described for wild ungulates, are unlikely to happen in intensive rangelands within the spatial scale adressed in the present study.

Because of these uncertainties and for reasons of simplicity we contrast the parent assumption of individual plant mortality beeing independent of vegetation cover with a constant amount of global feeding, which results in individual plant mortality linearly decreasing with vegetation cover. We expect constant global feeding to consolidate both stable states: an almost desertified landscape will be more resilient against recolonisation of single plants while the vegetated state will provide better protection to the plant individual. As a consequence, the reversibility of catastrophic shifts is expected to deteriorate.

@@

\begin{table}[t!]
\label{tab:models}
\caption{the models' substitutions for grazing,  $g$, in the mortality term $w_{ \left\{ +,0 \right\} }  = m_0 + g$.  }
\centering
\begin{tabular}{ccc}

@@

We applied these rules on a grid of 100 $\times$ 100 cells of 0.25m$^2$. Each timestep is defined as one year. The applicable transition probabilities for each single cell given in equations 1, 2, 3 and 4 are compared against uniform random numbers between 0 and 1 to determine if a transition occurs or if the cell remains unchanged. To initialize the grid with a standardized patchy vegetation structure, each model simulation was run for 500 years without any grazing (g = 0) using an initial environmental value $b_ini \in \left\{ 0.25, 3, 0.8, 0.9 \right\}$ before grazing is taking action. %(Note: I want to complete this to a highly resoluted gradient, to investigate the unstable equilibria). 
Starting from the preconditioned grids, the four grazing models (Table 1) with varying grazing intensity ($g \in \left{ 0, 0.025, 0.05, 0.075 , 0.10 \right}$) were run over a gradient of environmental quality ($b \in \left{ 0.2, ... , 1.0 \right}$) for 1.500 timesteps.

@@

To investigate the bistability properties of the vegetation cover, we simulate a gradient in environmental quality b ranging from 0.2 to 1 with a step-width of 0.025. The models of this study inherit all parameters from the parent model in K\'efi et al. 2007a ($r = 0.01$, $f = 0.9$, $\delta = 0.1$, $c = 0.2$, $d = 0.1$), except the ones regarding mortality. We assume intrinsic mortality m0 for all models to be 0.5, re
ecting an average individual lifespan of 20 years, if no additional mortality due to grazing is taking effect. We investigate, how the stability properties and patch size distributions change over a grazing gradient.

>>>>

\begin{figure}[!t]
	\centering
		\includegraphics{figures/fig1.pdf}
	\caption{a) Vegetation in arid ecosystems grows in patches. The associated growth provides enhanced local growth conditions as well as protection against grazing. With decreasing environmental quality, vegetation declines and the landscape becomes more fragmented (mid column), leading to desertification (right column). b) Schematic representation of plant patches. The direct neighborhood of plants provides shelter, retention of organic matter and water below and atop the soil surface. Spots remote from vegetation suffer from erosion and evaporation. c) The cellular automata model used in this study therefore distinguishes three states: occupied by vegetation (black), empty but fertile (grey) and degraded (white). d) Cumulative patch size distributions described by power law models: with decreasing environmental quality, the number of large patches declines, turning the power law distribution from up-bent (left) to down-bent (mid). Degraded landscapes have lost all the larger patches and are not to be described by a power law (right). }
	\label{fig:fig1}
\end{figure}

For arid ecosystems, a major mechanism behind this bistability was found to be local facilitation of plant individuals (REF?). The presence of a plant provides its direct local environment with abilities of water retention, organic matter accumulation and protection against external stressors (REF?). Opposingly, spots remote from any vegetation suffer most from erosion and evaporation, leading to locally degraded soils. As a consequence, the vegetation grows in patches (Aguiar and Sala 1999; Fig 1b).  When implemented in a simple spatially explicit model, this positive neighborhood effect provides true bistability on the landscape scale (K\'efi et al 2007a,b; Fig 1c). It therefore can be utilised to investigate the suitability of spatio-temporal metrics to indicate the proximity of a catastrophic shift, such as cumulative patch size distributions or critical slowing down.

This parent model was used before to investigate the consequences of increased plant mortality (K\'efi et al 2007a). It's simple formulation of additional plant stress can be interpreted as the effects of increased rangeland utilisation, corresponding to the grazing intensity by large herbivores, such as goat and sheep (K\'efi et al 2007a). However, the model has two features that must be considered unrealistic: the individual risk to be harmed by grazing (1) did not vary with total vegetation cover, and (2) it was defined globally and ignored the protective patch structure of the vegetation (Fig. 2a).

\begin{figure}[!tp]
	\centering
		\includegraphics{figures/fig2.pdf}
	\caption{Grazing affects the mortality of individual plants and the patch structure emerging in the landscape. The four models compared in this study use different assumptions on the quantity and spatial differentiation of a particular grazing pressure. Visualized in the left column are the plants dying due to grazing (red). In each model, plants suffer different individual risk, sometimes depending on the number of direct local neighbors (estimates for plants with 4, 2, or 0 neighbors).  a) The \textbf{parent model} as formulated in Kéfi et al 2008 assumes that grazing affects a proportion of living plants, i.e. individual risk is independent of global vegetation cover and shared evenly among all plants. b) The \textbf{livestock model} assumes a constant number of plants dying, regardless of the number of plants available. The individual risk is increasing when vegetation cover is low. c) Assuming \textbf{associative protection} leaves plants in the patch center unaffected by grazing (4), where­as plants at the patch border (2) or growing isolated (0) are suffering more from grazing. Particularly at high vegetation cover, the proportional grazing concentrates on those plants. d) The \textbf{combined model} integrates both assumptions leading to highest individual risk at low and high vegetation cover for isolated plants.  }
	\label{fig:fig2}
\end{figure}

Both assumptions are justified by the need for model simplicity, but disallow a translation of the model parameters to real livestock rates. Increasing individual plant mortality by factor two does not correspond to twice the number of animals per hectar: The proportional character of global plant mortality causes the total pressure on vegetation to decrease when intensive grazing reduces total vegetation cover (Fig 1a).

The relationship between livestock rates and the stability of vegetation cover is a crucial information for management decisions. The present study aims at increasing the realism and applicability of the model by applying two extensions that enable simulations of a grazing gradient on drylands.

%Grazing depends on vegetation cover
The first extension adds a contrasting assumption on the dependence on the total vegetation cover to the parent model. The prior model defined grazing as a constant addition to the risk of a plant to die. On the population level, the number of plants dying because of grazing decreased with a loss in vegetation cover (K\'efi et al. 2007, Fig. 2a). However, grazers forage in the landscape and consume to fulfill their daily demands. In other words, the number of plants suffering from grazing should be rather constant with vegetation cover. As a consequence, at low vegetation cover the grazing risk on individual plants will be high, whereas when vegetation cover is high, risk will be shared among many plants (Fig 2b).
In reality, the total feeding of large herbivores in the landscape is not entirely independent of vegetation cover but follows a non-linear saturating curve (type-II functional response, Holling 1959, Spalinger and Hobbs 1992). Thus, particularly at low vegetation cover, grazing is limited by the search success of the grazer, which is resource-density dependent. In rangeland systems, however, the functional response at low vegetation cover depends on factors on a larger scale: the animals move between landscapes motivated by internal decision patterns or their density is modulated by human management decisions (Senft 1989). We assume therefore, that smooth type-II functional responses, as described for wild ungulates, are unlikely to happen in intensive rangelands within the spatial scale adressed in the present study.

Because of these uncertainties and for reasons of simplicity we contrast the parent assumption of individual plant mortality beeing independent of vegetation cover with a constant amount of global feeding, which results in individual plant mortality increasing with vegetation cover (Fig 2b, right panel). We expect constant global feeding to consolidate both stable states: an almost desertified landscape will be more resilient against recolonisation of single plants while the vegetated state will provide better protection to the plant individual. As a consequence, the reversibility of catastrophic shifts is expected to deteriorate.

@@

\begin{table}[!th]
\label{tab:models}
\caption{the models' substitutions for grazing,  $g$, in the mortality term $w_{ \left\{ +,0 \right\} }  = m_0 + g$.  }
\centering
\begin{tabular}{ccc}

@@

We applied these rules on a grid of 100 $\times$ 100 cells of 0.25m$^2$. Each timestep is defined as one year. The applicable transition probabilities for each single cell given in equations 1, 2, 3 and 4 are compared against uniform random numbers between 0 and 1 to determine if a transition occurs or if the cell remains unchanged. To initialize the grid with a standardized patchy vegetation structure, each model simulation was run for 500 years without any grazing (g = 0) using an initial environmental value $ b_ini \in \left\{ 0.25, 3, 0.8, 0.9 \right\} $ before grazing is taking action. %(Note: I want to complete this to a highly resoluted gradient, to investigate the unstable equilibria). 
Starting from the preconditioned grids, the four grazing models (Table 1) with varying grazing intensity ($ g \in \left\{ 0, 0.025, 0.05, 0.075, 0.10 \right\} $) were run over a gradient of environmental quality ($ b \in \left\{ 0.2, ... , 1.0 \right\} $) for 1.500 timesteps.

@@

To investigate the bistability properties of the vegetation cover, we simulate a high resolution gradient in environmental quality $b$ ranging from 0.2 to 1 with varying step-width between 0.0025 and 0.025. The models of this study inherit all parameters from the parent model in K\'efi et al. 2007a ($r = 0.01$, $f = 0.9$, $\delta = 0.1$, $c = 0.2$, $d = 0.1$), while substituting mortality with Eq. 4 and Table 1. We assume intrinsic mortality $m_0$ for all models to be 0.05, reflecting an average individual lifespan of 20 years, if no additional mortality due to grazing is taking effect.  
Further, we run all models over a linear grazing gradient with $g$ taking 0, 0.025, 0.050, 0.075, 0.100.
For the parent model, this produces average individual lifespans of 20.0, 13.3, 10.0, 8.0, 6.7 years, respectively.

\subsection{Cumulative patch--size distributions}
We investigate qualitatively, how the patch size distributions change with environmental quality at the different levels of grazing. Therefore, we assigned each simulation run to one of the following five cases: (1) fully vegetated, if the vegetation cover was significant (\> 2 \% ) and aggregated into one large spanning cluster (ignoring additional patches up to a size of 2 cells); (2) up-bent power law, if the cumulative patch--size distribution was best described by a power law with a lower limit defined by a single large spanning cluster; (3) power law, if the cumulative patch--size distribution  was best described by a power-law; (4) down-bent power law, if the cumulative patch--size distribution was best described by a truncated power law; (5) degraded, if the vegetation cover was insignificant (\< 2\%).

We compare the parameter space occupied by those cases. Additionally, we compare the exponent of the power law for the cases 2--4.
